### 2016-02-03 hans/bayes-tardis-luminosity-prediction.tex
[hans] Add section on approximations
* synctex support, requires pdflatex
* bibtex references

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Bayesian prediction of luminosity distribution based on one simulation run:
% An example from TARDIS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[11pt]{article}
\usepackage[a4paper, left=20mm, right=20mm, top=25mm, bottom=25mm]{geometry}
\usepackage[numbers]{natbib}
%\usepackage{hyperref}

@@

\DeclareMathOperator{\GammaDist}{Gamma}
\newcommand{\refeq}[1]{Eq.~(\ref{#1})}
\newcommand{\lmax}{\ell_{\rm max}}

@@

The Gamma distribution is with shape parameter $\alpha$ and rate
parameter $\beta$ is defined for $x \geq 0$ as
\begin{align}
  \lleq{gma}
  \GammaDist(x | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x} .
\end{align}
$\GammaDist(x \cond \alpha, \beta)$ is defined for $x \in [0, \infty]$
whereas the luminosity of real packets is bounded from above and has a
peak away from the bound. In order to describe the luminosity
distribution of real packets through $\GammaDist(\alpha, \beta)$, we
have to transform the luminosities. To this end, we first take the
largest luminosity from the entire frequency range and increase it
slightly to avoid numerical difficulties at the end point
\begin{align}
  \lleq{gme}
  \lmax \equiv (1+\epsilon) \times \max_j \ell_j,
\end{align}
where $ \epsilon=10^{-6}$.
We assume the total number of samples across all bins is large, hence
the effect of considering $\lmax$ a constant independent of the data
is negligible. We then transform the samples as
\begin{align}
  \lleq{gmf}
  x_j \equiv 1- \frac{\ell_j}{\lmax} > 0
\end{align}
and similarly for $L \to X$, then we infer $P(X \cond n, \bmx,
m)$. Only in the end, we transform back to a probability density in
$L$ via
\begin{align}
  \lleq{gmg}
  p(L  \cond n, \bml, m) = p(X  \cond n, \bmx, m) \frac{1}{\lmax}
\end{align}
While not obvious at first, a major numerical simplification arises
because of the $L \to X$ transformation. Rewriting the master
equation~(\ref{pmp}) formally as
\begin{align}
  \lleq{gmh}
  p(L  \cond n, \bml, m) = \sum_{N} p(N \cond n) p(L \cond N, \bml, m)
\end{align}
we see that the $ p(L \cond n, \bml, m)$ is a mixture density. If the
distribution of a single luminosity, $p(L \cond N=1, \bml, m)$ is
narrow in the sense that the standard deviation relative to the mean
is at the percent level, then the mixture density has one mode for
each $N$ and all modes are clearly separated. The transformation $L
\to x$ effects a strong overlap between adjacent $p(X \cond N, \bmx,
m)$ and leads to a unimodal $p(X \cond n, \bmx, m)$.
\fred{Show a plot: from hedgehog to single peak}

For our purposes, the most useful property of the Gamma
distribution is infinite divisibility; i.e., the sum of $N$ Gamma
variates is again a Gamma variate
\begin{align}
  \lleq{gmb}
  X_i \sim \GammaDist(\alpha, \beta) \Rightarrow \sum_{i=1}^N X_i \sim \GammaDist(N \alpha, \beta).
\end{align}
With $\bmphi = (\alpha, \beta)$, \refeq{gmb} is equivalent to
\begin{align}
  \lleq{gmc}
 p(X \cond N, \bmphi) &= \int d\bmX\,\delta(X - \textstyle\sum_j X_j)
  \,p(\bmX\cond N,\bmphi) = \GammaDist(X \cond N \alpha, \beta)
\end{align}
and this implies that the master equation~(\ref{pmp}) considerably simplifies as
\begin{align}
  \lleq{gmd}
    p(X\cond n,\bmx,m)
  = \frac{1}{p(\bmx\cond m)}
  \sum_N p(N\cond n)\, \int d\bmphi\, p(X \cond N, \bmphi)\, p(\bmx\cond m,\bmphi)
  \, p(\bmphi).
\end{align}

To explicitly evaluate \refeq{gmd}, we have to overcome the remaining
two difficulties: the sum over $N$ and the integral over
$\bmphi$. Regarding the latter, there is a conjugate prior but
unfortunately the normalization constant can not be evaluated in
closed form. It is thus of no use for us, and we turn to a
noninformative prior. The authors of \cite{moala2013bayesian} present
an overview of possible choices and conclude that the point estimate
given by the mode of the posterior is nearly insensitive to the prior
choice for 30 samples. We require in addition that the posterior is
narrow enough so we group samples in bins to have at least $m=150$
samples in each bin. The following approximations can be used in
different regions of the parameter space spanned by $n$ and $m$.

\begin{enumerate}
  \item For $m$ large, the posterior $p(\bmphi \cond m, \bmx) =
  p(\bmx\cond m,\bmphi) \, p(\bmphi) / p(\bmx \cond m)$ approaches a
  multivariate Gaussian with variance tending to zero as $1/m$ in the
  limit $m \to \infty$.  \fred{check} We can then assume the
  parameters $\bmphi$ known and fix them at the posterior mode
  $\bmphi^*$; i.e., $p(\bmphi \cond m, \bmx) \to \delta(\bmphi -
  \bmphi^*)$.
  \item For $n$ large, the Negative Binomial $p(N \cond n)$ approaches
  a Gaussian, defining a region around $N=n$ with the dominant
  contribution to the sum over $N$ in \refeq{gmd}. The central limit
  theorem states that for large $N$, $P(X \cond N, \bmphi) \approx
  \mathcal{N}(X \cond N \, E_\bmphi[X_j], N \, V_\bmphi[X_j])$. For
  $m$ large enough, we could then plug in the sample-mean and
  sample-variance estimates and bypass the assumption of a Gamma
  distribution altogether.
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Prediction based on packet sum}

TODO: Hans. The calculations have already been done. The question is
whether we even want to bother with them since data in the form of
packet sums is less information than the vector of individual packets
as used above.
\fred{No, we don't want to bother.}

@@

In Figure~\ref{fig:beta} is a fit of the Beta distribution to 3000
samples from a TARDIS run with frequencies in the range $[1.367,
1.435] \times 10^{38}$ Hz. A characteristic feature seen in many such
similar fits is that the Beta distribution does not give a perfect
fit: it undershoots around the mode and overshoots around
$x=0.96$. The reason may be that the Beta itself is not flexible
enough to describe the data, or the distribution is simply not
constant in frequency, so we should choose smaller bins or less
samples. We don't expect 3000 samples in every bin, so what is the
minimum number of samples needed to work robustly? The first numerical
challenge is how to normalize the luminosities, I take the maximum
across all samples (not just one bin) and multiply by $10^{-6}$ to
avoid $x=1$ for any sample. For too few samples, the fit returns a pdf
that peaks at the boundary $x=1$, this is not what we want. From trial
and error, 300 samples is a solid minimum to get a robust fit if
parameters are estimated via maximum likelihood. The method-of-moment
estimator has the advantage of not requiring an optimization algorithm
but for some bins it gives a solution peaking at $x=1$ even for 800
samples so it's not robust.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.48\textwidth]{beta-300}
  \includegraphics[width=0.48\textwidth]{beta-3000}
  \caption{Beta distribution fit. The samples are binned in the
    histogram. The blue solid curve is a kernel-density estimate, the
    red dashed curve is the Beta distribution. left: 300 samples,
    right: 3000 samples}
  \label{fig:beta}
\end{figure}

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Bayesian prediction of luminosity distribution based on one simulation run:
% An example from TARDIS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[11pt]{article}
\usepackage[a4paper, left=20mm, right=20mm, top=25mm, bottom=25mm]{geometry}
\usepackage[numbers]{natbib}
%\usepackage{hyperref}
\synctex=1

@@

\DeclareMathOperator{\GammaDist}{Gamma}
\newcommand{\refeq}[1]{Eq.~(\ref{#1})}
\newcommand{\reffig}[1]{Fig.~\ref{fig:#1}}
\newcommand{\lmax}{\ell_{\rm max}}

@@

The Gamma distribution  with shape parameter $\alpha$ and rate
parameter $\beta$ is defined
\begin{align}
  \lleq{gma}
  \GammaDist(x | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x} .
\end{align}
The support of $\GammaDist(x \cond \alpha, \beta)$ is $[0, \infty]$
whereas the luminosity of real packets is bounded from above and has a
peak close to but not at the boundary. In order to describe the luminosity
distribution of real packets through $\GammaDist(\alpha, \beta)$, we
have to transform the luminosities. To this end, we first take the
largest luminosity from the entire frequency range and increase it
slightly to avoid numerical difficulties at the end point
\begin{align}
  \lleq{gme}
  \lmax \equiv (1+\epsilon) \times \max_j \ell_j,
\end{align}
where $ \epsilon=10^{-6}$.
We assume the total number of samples across all bins is large, hence
the effect of considering $\lmax$ a constant independent of the data
is negligible. We then transform the samples as
\begin{align}
  \lleq{gmf}
  x_j \equiv 1- \frac{\ell_j}{\lmax} > 0
\end{align}
and similarly for $L \to X$, then we infer $P(X \cond n, \bmx,
m)$. Only in the end, we transform back to a probability density in
$L$ via
\begin{align}
  \lleq{gmg}
  p(L  \cond n, \bml, m) = p(X  \cond n, \bmx, m) \frac{1}{\lmax}
\end{align}
We observe good agreement between the best fit of a Gamma distribution
to TARDIS samples; an example is shown shown in \reffig{gamma}.  While
not obvious at first, a major numerical simplification arises because
of the $L \to X$ transformation. Rewriting the master
equation~(\ref{pmp}) formally as
\begin{align}
  \lleq{gmh}
  p(L  \cond n, \bml, m) = \sum_{N} p(N \cond n) p(L \cond N, \bml, m)
\end{align}
we see that the right-hand side is a mixture density. In our
application, the distribution of a single luminosity, $p(L \cond N=1,
\bml, m)$ is narrow in the sense that the standard deviation relative
to the mean is at the percent level. This implies that $p(L \cond n,
\bml, m)$ has one mode from each $N$ and all modes are clearly
separated. The transformation $L \to x$ effects a strong overlap
between adjacent $p(X \cond N, \bmx, m)$ leading to a unimodal $p(X
\cond n, \bmx, m)$.  \fred{Show plot: from hedgehog to single peak}

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.48\textwidth]{gamma}
  \caption{Gamma distribution fit. }
  \label{fig:gamma}
\end{figure}

For our purposes, the most useful property of the Gamma
distribution is infinite divisibility; i.e., the sum of $N$ Gamma
variates is again a Gamma variate
\begin{align}
  \lleq{gmb}
  X_j \sim \GammaDist(\alpha, \beta) \Rightarrow \sum_{j=1}^N X_j \sim \GammaDist(N \alpha, \beta).
\end{align}
With
\begin{align}
  \lleq{gmi}
  \bmphi &= (\alpha, \beta),\\
  p(X_j \cond \bmphi) &=
  \GammaDist(X_j \cond \alpha, \beta),\\
  p(\bmX \cond N, \bmphi) &= \prod_{j=1}^N p(X_j \cond \phi)
\end{align}
 \refeq{gmb} is equivalent to
\begin{align}
  \lleq{gmc}
 p(X \cond N, \bmphi) &= \int d\bmX\,\delta(X - \textstyle\sum_j X_j)
  \,p(\bmX\cond N,\bmphi) = \GammaDist(X \cond N \alpha, \beta)
\end{align}
and this implies that the master equation~(\ref{pmp}) considerably simplifies as
\begin{align}
  \lleq{gmd}
    p(X\cond n,\bmx,m)
  = \frac{1}{p(\bmx\cond m)}
  \sum_N p(N\cond n)\, \int d\bmphi\, p(X \cond N, \bmphi)\, p(\bmx\cond m,\bmphi)
  \, p(\bmphi).
\end{align}

\subsection*{Approximations}

To explicitly evaluate \refeq{gmd}, we have to overcome the remaining
two difficulties: the sum over $N$ and the integral over
$\bmphi$. Regarding the latter, there is a conjugate prior but
unfortunately the normalization constant cannot be evaluated in closed
form. It is thus of no use for us, and we turn to a noninformative
prior. \fred{Be explicit about prior choice} The authors of
\cite{moala2013bayesian} present an overview of possible choices and
conclude that the point estimate given by the mode of the posterior is
nearly insensitive to the prior choice for 30 samples. We require in
addition that the posterior is narrow enough so we group samples in
bins to have at least $m=150$ samples in each bin. The following
approximations can be used in different regions of the parameter space
spanned by $n$ and $m$.

\begin{mnumerate}
  \item \label{large-m} For $m$ large, the posterior $p(\bmphi \cond m, \bmx) =
  p(\bmx\cond m,\bmphi) \, p(\bmphi) / p(\bmx \cond m)$ approaches a
  multivariate Gaussian. We can then compute the various moments
  $E_\bmphi[f(x)]$ with $f(x)=1,x,x^2$ by Laplace's method to good
  accuracy.
  \item \label{very-large-m} In the limit $m \to \infty$, the covariance of the posterior
  $p(\bmphi \cond m, \bmx)$ tends to zero as $1/m$ and we can then fix
  the parameters $\bmphi$ at the posterior mode $\bmphi^*$; i.e.,
  $p(\bmphi \cond m, \bmx) \to \delta(\bmphi - \bmphi^*)$.
  \item \label{large-n} For $n$ large, the cumulative of the Negative
  Binomial $p(N \cond n)$ can be approximated by a Gaussian, defining
  a region around $N=n$ with the dominant contribution to the sum over
  $N$ in \refeq{gmd}. The contribution around $N=0$ is negligible, and
  we can replace $\sum_n \to \int dN$. The density $p(N \cond n) P(X
  \cond N,m, \bmx)$ is close to a Gaussian and the integral can be
  accurately approximated by Laplace's method which requires the
  Hessian and the mode. We benefit from partial error cancelation
  because we apply Laplace's method both in the numerator and
  denominator~\cite{tierney_accurate_1986} as each moment has to be
  normalized by the evidence $p(\bmx \cond m)$. The former is
  available in closed form and the latter has to be determined
  numerically. Fortunately, Newton's method leads to a solution in
  only a few steps~\cite{minka2002estimating}. \fred{The reference is
    for max. likelihood of a $\GammaDist(\cdot)$ but I'm sure the
    performance is similar for max. posterior and the first moments}
  \item \label{central-limit} The central limit theorem states that for large $N$,
  $P(X \cond N, \bmphi) \approx \mathcal{N}(X \cond N \,
  E_\bmphi[X_j], N \, V_\bmphi[X_j])$. For $m$ large enough, we could
  then plug in the sample-mean and sample-variance estimates and
  bypass the assumption of a Gamma distribution altogether.
  \item \label{large-n-large-m} For large $n$ and $m$, $ p(X\cond
  n,\bmx,m)$ can be approximated by a Gaussian whose mean and variance
  are given by \refeq{mgo} and \refeq{mgr} in terms of
  $\bmphi$-expectation values. The latter are integrals with no known
  closed-form solutions but we can easily and accurately approximate
  them with Laplace's method.
\end{mnumerate}
To simplify, we assume that $m$ is large. In our application, we may
have to combine several frequency bins to achieve $m \gtrsim 150$ for
the purpose of infering $\bmphi$. This is equivalent to assuming that
the distribution changes only slowly with the frequency, a condition
that we verified for TARDIS. It is thus justified to use
approximations \ref{large-m} or even \ref{very-large-m}.

Numerically, the most challenging case is small $n$ because the sum
over $N$ in \refeq{gmd} has to be computed explicitly starting at
$N=1$ until contributions become negligible. Since $p(N \cond n)$
falls off quickly, it is sufficient in our application to do this
until $n \lesssim 20-30$. Beyond that, we use approximations
\ref{large-n}--\ref{large-n-large-m}.

\subsection*{Comparison of approximations}

@@

In \reffig{beta} is a fit of the Beta distribution to 3000
samples from a TARDIS run with frequencies in the range $[1.367,
1.435] \times 10^{38}$ Hz. A characteristic feature seen in many such
similar fits is that the Beta distribution does not give a perfect
fit: it undershoots around the mode and overshoots around
$x=0.96$. The reason may be that the Beta itself is not flexible
enough to describe the data, or the distribution is simply not
constant in frequency, so we should choose smaller bins or less
samples. We don't expect 3000 samples in every bin, so what is the
minimum number of samples needed to work robustly? The first numerical
challenge is how to normalize the luminosities, I take the maximum
across all samples (not just one bin) and multiply by $10^{-6}$ to
avoid $x=1$ for any sample. For too few samples, the fit returns a pdf
that peaks at the boundary $x=1$, this is not what we want. From trial
and error, 300 samples is a solid minimum to get a robust fit if
parameters are estimated via maximum likelihood. The method-of-moment
estimator has the advantage of not requiring an optimization algorithm
but for some bins it gives a solution peaking at $x=1$ even for 800
samples so it's not robust.

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.48\textwidth]{beta-300}
  \includegraphics[width=0.48\textwidth]{beta-3000}
  \caption{Beta distribution fit. The samples are binned in the
    histogram. The blue solid curve is a kernel-density estimate, the
    red dashed curve is the Beta distribution. left: 300 samples,
    right: 3000 samples}
  \label{fig:beta}
\end{figure}

@@

% Local Variables:
% compile-command:"rubber --pdf -W refs -W misc -S bayes-tardis-luminosity-prediction"
% End:
